---
title: "Tutorial 5"
toc: false
---

Tutorial 5 was the mini coding project 2 assessment task. You can find your mark and specific comments on where you lost marks on canvas. 

## Solutions
Sample solutions below. Note that I accepted any valid code that gives the expected result, and so your answer may not be in this pdf. There are so many ways to do things in R I can't possibly include them all!

{{< pdf "resources/solution/mini_coding_project_2_solutions.pdf" >}}

## Extra comments on poorly-answered questions
### Question 5

**Question.** Testing whether `letters` and `pi` have the same atomic type (class)

The key word in this question is **testing**, which means you **need to do a logical test**, like so:

```{r}
class(letters) == class(pi)
```

Many people lost 0.5 marks by only finding the class of each variable, but not using `==` to test for their equality 

### Question 6

**Question.** I want to find the mean (average) of the numbers 1, 2, 3, 4 and 5 using R. I tried using the following code:

```{r}
mean(1, 2, 3, 4, 5) 
```

We know the mean of 1,2,3,4,5 should be 3, but running this code in
RStudio gives the wrong answer.

a.  Explain why (1 mark)
b.  In the code chunk below, correct the mistake (1 mark)

Here, the problem is that the numbers 1-5 are not in a vector, they are simply provided as arguments to the `mean()` function. For a), I accepted any answer along the lines of 'they should be in a vector'.
For b), you just needed to pop them into a vector like `mean(c(1,2,3,4,5))` or `mean(1:5)`.

### Question 7

**Question.** Inside this assessment's R project, there is a folder called 'data_for_part_b', that contains a file called 'melanoma_dataset.tsv'. Please write code to read this data into R, so that you can use it to answer the subsequent questions.

Here, to get full marks you needed to **read in the data file** and **assign it to a variable** so that you could use it in the subsequent questions. Many people had perfectly correct `read_delim()` function calls, but forgot to assign it to a variable!

Another common problem was forgetting to put quotation marks `""` or `''` around the path, i.e. `read_delim(data_for_part_b/melanoma_dataset.tsv)` instead of `read_delim("data_for_part_b/melanoma_dataset.tsv")`.

Finally, some people provided a path to the folder, not to the file. For example `"~/Desktop/mini_coding_project_1/data_for_part_b"` instead of `"~/Desktop/mini_coding_project_1/data_for_part_b/melanoma_dataset.tsv"`
`read_delim()` needs a path to the file you want to read in, so your path should always end in something.csv or something.tsv! 

**Please revise this before the next mini coding project, it will appear again**

### Question 8c

```{r}
#| echo: false
library(tidyverse)
melanoma_data <- tibble(boot::melanoma)
```

**Question.** 
The researchers are interested specifically in the melanoma of patients born after 1920. However, they have only recorded the age of the patients at diagnosis (`age` column) and the year in which they were diagnosed (`year` column).

a.  Looking at the `melanoma_data` dataframe, **explain in words** how
    you could find birth year from this data (1 mark)

For part a, most of the issues revolved around what 'age' should be subtracted from. Reading the question carefully, it is the age **at diagnosis** not necessarily their current age. In fact, this data is over 50 years old, so I don't think many of the patients are alive anymore! Anyway, you needed to do the 'year' column, minus the 'age' column.

b.  Write R code to create a column called `birth_year` in
    `melanoma_data`. Ensure that the result is saved as a variable (2
    marks)

You could either do this with base R (workshop 1):

```{r}
melanoma_data$birth_year <- melanoma_data$year - melanoma_data$age
```

Or using dplyr (workshop 2):

```{r}
melanoma_data <- melanoma_data %>%
  mutate(birth_year = year - age)
```

If using dplyr, **you needed to remember to assign the result** to something (many people forgot this). If not, you would get an error for Q10c).

c.  Using your column from part b, how many patients were born after
    1920? (1 mark)
    
You could either do this with base R (workshop 1):

```{r}
sum(melanoma_data$birth_year > 1920)
```

Or using dplyr (workshop 2):

```{r}
melanoma_data %>%
  filter(birth_year > 1920)
```
    
Common errors included using mutate to create a new column that says if a patient was born after 1920 (but not counting this up), or accidentally filtering on the original 'year' column not the 'birth_year' column just created (in this case, since the data was collected during 1962-1977, all rows pass your filter and you get an answer of 205 instead of the correct answer of 80)
