---
title: "Tutorial 4"
toc: false
---

In this tutorial we learned about the `esquisse` R package for making graphs with `ggplot2` through an interactive graphical interface.

We first loaded in the `msleep` data from `ggplot2` to practice with, and revised our `dplyr` skills:

```{r}
# load tidyverse
library(tidyverse)

# we will work with the msleep dataset
data(msleep, package = "ggplot2")
msleep

# practice activity to revise workshop 3
# use the summarise() with grouping to find the average amount of sleep
# 'sleep_total' column for each of carnivores, omnivores and herbivores
msleep %>%
  summarise(
    av_sleep = mean(sleep_total),
    n = n(),
    .by = vore
  )
```

Then, we set up `esquisse` as follows:

```{r}
#| eval: false

# install package
install.packages("esquisse")

# instead of esquisse:: we could use library(esquisse)
# package_name::function() lets you use a function without loading 
# in the whole package (useful when you have many packages you want to use)
esquisse::esquisser(msleep)
```

You can find more information about esquisse [on their web page](https://dreamrs.github.io/esquisse/).

We then made a graph in the interactive interface, and copied the `ggplot2` code for our records. Here is my plot:

```{r}
# always make sure to copy the code 
# this is mine
ggplot(msleep) +
  aes(x = bodywt, y = brainwt, colour = vore) +
  geom_point() +
  scale_color_hue(direction = 1) +
  scale_x_continuous(trans = "log") +
  scale_y_continuous(trans = "log") +
  labs(
    x = "Body weight",
    y = "Brain Weight (kg)",
    title = "Is there a relationship between the size of the animal and its brain size?",
    color = "Diet"
  ) +
  theme_minimal() +
  theme(
    legend.text = element_text(size = 12L),
    legend.title = element_text(size = 12L)
  )
```

We will cover more about `ggplot2` code in workshop 4 next week!