---
title: "Tutorial 5"
toc: false
---

Tutorial 5 was the mini coding project 2 assessment task. You can find your mark and specific comments on where you lost marks on canvas. 

```{r}
library(tidyverse)
```

## Solutions
Sample solutions below. Note that I accepted any valid code that gives the expected result, and so your answer may not be in this pdf. There are so many ways to do things in R I can't possibly include them all!

{{< pdf "resources/solution/mini_coding_project_2_solutions.pdf" >}}

## Extra comments on poorly-answered questions
### Question 5

**Question.** I want to create a wide-format summary table showing the average viability for each combination of cell line and treatment, so I can know which cell line is the most drug-resistant. Write R code to create this table. Hint: you first need to find the average viability for each
cell line/treatment combination with a grouped `summarise()`, then use `pivot_wider()` to construct the table.

The hint described the two functions you needed to use, and most people were at least on the right track with this:

```{r}
#| eval: false
# first, do the summarise
cell_viability %>%
  summarise(
    av_viability = mean(viability_percent, na.rm = TRUE),
    .by = c(cell_line, treatment)) %>%
  # then, pivot_wider
  pivot_wider(
    # names_from is the new column names, which come from
    # the treatment column
    names_from = treatment,
    # values_from is what values belong in the table, so the
    # average viability. NOT the 'viability_percent' 
    values_from = av_viability
  )
```

One of the most common mistakes here was creating the average viability column during the `summarise()` and then trying to use the percent_viability column for the pivot. Another common mistake was using the wrong column entirely, and creating a table of the average cell count, rather than the average percentage viability

Some people didn't remove missing values (with either `na.omit()` or `na.rm = TRUE`), although I didn't deduct marks for this as it wasn't explicitly asked for in the question.

### Question 6

**Question.** Your task is to wrangle `plate_reader_export` from a wide format into a long format. What do you think a long-form version of this data would look like (i.e. what columns might it have)?

Here, you were being asked to describe the data after pivoting: anything that mentioned more rows and less columns was acceptable for 0.5 marks, and specifically anything that talked about the dosage columns all being put into one column was awarded the full 1 mark

**Question.** Write code to create the long-form version of this dataframe

Here you needed to use the `pivot_longer()` function as described in the question, specifically on the dose columns:

```{r}
#| eval: false
plate_reader_export %>%
  pivot_longer(
    # columns to pivot
    cols = starts_with("dose_"),
    # names of our new columns
    names_to = "dose_uM",
    values_to = "viability_percent"
  )
```

Some people tried to pivot on the wrong columns, the treatment and cell line columns are already in long format (wide format would be having a column for each cell line, or a column for each drug).

Another common error was naming the columns for the `names_to` and `values_to` arguments. These are **the names of new columns** and so **must not already exist in the data frame**. Several people used `"treatment"` for example, which won't work as it collides with the existing column called treatment.

### Question 7

**Question.** I would like to add the `cell_line_info` dataframe into the `cell_viability` dataframe.
a. Write code to join these two dataframes
b. Is there a difference between the result of joining `cell_line_info` to `cell_viability` or doing it the opposite way (`cell_viability` to `cell_line_info`)? In other words, does it matter which order we join the dataframes in?
c. Imagine that cell_line_infowas missing the HEK293 cell line. In this case, would it matter which order we join the dataframes in?

Overall, Q7a was well answered, most people who attempted it got full marks.

For Q7b and c however, the explanations were tricky! To be clear, the order you join in when using `left_join()` matters if, **for the column you are joining on**, there are values in one data frame without a match in the other. 

So in Q7b, because all cell lines appear in both `cell_viability` and `cell_line_info`, the resulting output will have the same number of rows and columns, regardless of the way you use `left_join()` to combine them. The only thing that will be different is the order of those rows and columns.

In Q7c, since HEK293 is removed from the `cell_line_info` dataframe, the order of joining is important, as if `cell_line_info` goes first, all HEK293 rows from `cell_viability` will be dropped, whereas if `cell_viability` is first, all HEK293 rows will remain, with NAs used to fill in the missing values.

### Question 8c

**Question.** 
I want to highlight in my presentation the % viability of HEK293 cells at a dose of 100 uM in Drug B. Write R code to extract this value out of dose_summary.

Here, you needed to use filter, and convert the text into logical tests. I think maybe the difficulty with this question was due to the fact that the wording is a bit different from how `filter()` has been presented in the past.

```{r}
#| eval: false
dose_summary %>%
  filter(treatment == "DrugB") %>%
  filter(cell_line == "HEK293") %>%
  filter(dose_uM == 100)
```

Many people lost marks for not including all three logical tests, or simply using other unrelated functions like `summarise()` and `select()`