---
title: "Further Reading"
toc: false
---

## The difference between the base R pipe and magrittr pipe {#sec-fr-pipes}

In @sec-pipes, we introduced the `magrittr` pipe `%>%` and also mentioned the 
base R alternative `|>`. This further reading will discuss the differences 
between them and why you might choose one over the other.

When looking at older R code, you'll tend to see `%>%`, as the `|>` pipe was 
only introduced relatively recently (2023). This is why I still use `%>%` most
of the time, as it's what I learnt way back when I started with R!

For most purposes, the two pipes function the same, passing data from the 
left-hand side to the right-hand side. However, there are a couple of differences:

### Placeholders
In @sec-pipes, we learnt that a pipe passes data into the **first argument** of
the function on the right-hand side. While this is the default behaviour, it's 
actually possible to change this using placeholders `.` for the magrittr pipe
and `_` for the base R pipe. 

For example, let's say we have the function f(x, y), which we want to run with
the values x = 1 and y = 2:

```{r}
#| eval: false
f(1, 2)
```

Let's assume now that the value of x (1) comes from some other code/function. We
can pipe it into `f(x,y)` like so:

```{r}
#| eval: false

# with magrittr pipe
1 %>% f(2)

# with base R pipe
1 |> f(2)
```

But what if we wanted to pipe in the value 2 instead? We can achieve this using
a placeholder:

```{r}
#| eval: false

# with magrittr pipe
2 %>% f(1, .)

# with base R pipe
2 |> f(1, _)
```

### Dropping parentheses
In @sec-workshop02, we always used brackets when writing out our functions, 
even if they had no arguments, for example:

```{r}
# we need to load tidyverse before using the pipe
library(tidyverse)

iris %>%
  # still put the brackets even though they're empty
  head()
```

However, it's actually possible with the `%>%` pipe to skip the brackets:

```{r}
iris %>%
  # as long as there's no arguments, don't need brackets with %>%
  head
```

Personally, I still like to use the brackets to remind myself that these are
functions, not variables. But some people like to leave them out. This isn't 
possible with `|>`:
```{r}
#| error: true
iris |>
  # as long as there's no arguments, don't need brackets with %>%
  head
```

We can see that this gives us an error.

### Use in packages
The `|>` pipe is in base R, that is it comes with every copy of R, without you
needing to install a package. This can be advantageous, particularly when doing
software development as your software will not depend on as many other packages
(by using `%>%` you need to depend on the `magrittr` package). As we're not 
doing any software development in this course you won't need to worry about it, 
but it could be important to keep in mind for your future, if you ever want to 
make your own R package.