---
title: "Further Reading"
filters:
  - naquiz
toc: false
---

## The rest of the join family {#sec-manyjoins}

`left_join()` covers most basic joining tasks, but dplyr also provides many 
other join functions that can be useful. They all share the same basic 
syntax as `left_join()`, taking two data frames and a `by` argument, but differ 
in their output.

To explore these different join functions, we'll use two new data frames: a 
sample sheet, and a QC report showing the number of reads from each sample. There 
are a few deliberate mismatches between them, which will help us understand how 
the different joins behave:

```{r}
# first load the packages
library(tidyverse)

sample_sheet <- tibble(
  sample_id = c("K01", "K02", "K03", "K04", "K05", "K06"),
  ko_gene   = c("NINJ1", "NINJ1", "control", "NINJ1", "control", "NINJ1"),
  operator  = c("KP", "KP", "KP", "JL", "JL", "JL")
)

qc_report <- tibble(
  sample_id  = c("K01", "K02", "K04", "K06", "K07"),
  qc_status  = c("pass", "pass", "pass", "pass", "pass"),
  total_reads = c(48000, 51000, 39500, 44200, 60000)
)
```

Notice that `K03` and `K05` are in the sample sheet but never made it into 
the QC report (maybe their sequencing failed, or hasn't been completed yet), and
`K07` appears in the QC report but isn't in the sample sheet at all (maybe a 
typo, or a sample from a different batch got mixed in).

### Mutating joins

These joins all add columns from one data frame to another, but differ in which 
rows survive.

**`right_join()`** is the mirror image of `left_join()` — it keeps all rows 
from the *right* data frame instead of the left:

```{r}
right_join(
    sample_sheet,
    qc_report,
    by = "sample_id")
```

Actually, `right_join(x, y)` gives the exact same result as `left_join(y, x)`.
So, most people just use `left_join()` instead.

**`inner_join()`** only keeps rows that have a match in *both* data frames:

```{r}
inner_join(
    sample_sheet,
    qc_report,
    by = "sample_id")
```

This returns only `K01`, `K02`, `K04`, and `K06` — the four samples 
present in both data frames. `K03` and `K05` (missing from `qc_report`) and 
`K07` (missing from `sample_sheet`) are all dropped. This is useful when you 
specifically want to analyse only observations that have complete 
information from both data frames

**`full_join()`** keeps every row from *both* data frames, filling in `NA` where 
there's no match on either side:

```{r}
full_join(
    sample_sheet,
    qc_report,
    by = "sample_id")
```

This returns all 7 unique samples: the 4 that matched, plus `K03` and `K05` 
(with `NA`s for QC columns) and `K07` (with `NA`s for sample sheet columns).

### Filtering joins

Unlike the mutating joins above, filtering joins never add new columns, and 
instead keep or remove rows from the left data frame based on whether they match
the right data frame. This makes them useful for checking or filtering your data
rather than combining it.

**`semi_join()`** keeps rows in the left data frame that **do have a match** in 
the right data frame (but doesn't add any of the right data frame's columns):

```{r}
semi_join(
    sample_sheet,
    qc_report,
    by = "sample_id")
```

This returns `NB01`, `NB02`, `NB04`, and `NB06`, but notice the result only 
contains columns from `sample_sheet` (`sample_id`, `antigen`, `operator`). None 
of the columns from `qc_report` (`qc_status`, `total_reads`) are added.

**`anti_join()`** does the opposite: it keeps rows in the left data frame that 
**don't have a match** in the right data frame:

```{r}
anti_join(
    sample_sheet,
    qc_report,
    by = "sample_id")
```

This returns `K03` and `K05`, the two samples missing from the QC report. It 
goes to show why `anti_join()` is a great debugging tool: if you expect every 
sample in `sample_sheet` to also appear in `qc_report`, running `anti_join()` 
will immediately flag any samples that are missing.

### A quick summary

| Function | Keeps rows from | Adds columns? |
|---|---|---|
| `left_join()` | left data frame | Yes |
| `right_join()` | right data frame | Yes |
| `inner_join()` | rows matching in both | Yes |
| `full_join()` | all rows from both | Yes |
| `semi_join()` | left rows matching right | No |
| `anti_join()` | left rows *not* matching right | No |

For the full documentation and more examples, see the 
[dplyr two-table verbs vignette](https://dplyr.tidyverse.org/articles/two-table.html).
